Galaxy Zoo: the dependence of the star formation-stellar mass relation on spiral disk morphology

### 2015-02-04 response_gzmainsequence.txt

We thank the reviewer for the thoughtful read and suggestions for improving this manuscript. We have addressed all of the issues, primarily centered on analyzing a smaller sample of galaxies due to concerns about completeness of star formation rates at lower masses. The main scientific conclusion of this paper --- namely that changes in disk morphology do not affect the position of a galaxy on the star-forming main sequence --- have not changed as a result of this new analysis.

A more detailed response and list of the changes is attached below. We hope the manuscript is now ready for publication.

Reviewer's Comments:

Comments to the Author
The paper « Galaxy Zoo: the dependence of the star formation-stellar mass relation on spiral disk morphology » compares the positions of star-forming galaxies of various disk morphologies to the best fitting line of the SFR-M* relation. This study presents an interesting discussion of e.g. the role of multiple arms in disk galaxies but suffers from a major problem that strongly limits the interpretation. All the analysis relies on the comparison of a sub-sample of the galaxies with a reference trend shown in the figures with a black solid line. It is the definition of this line, more precisely its slope and normalization that seems probmematic to me.

Although this is not clearly stated in the text, this line was apparently obtained by fitting the whole sample of star-forming galaxies with no restriction in stellar mass or SFR. I believe that such fit can be misleading for the following reason. There is necessarily a completeness limit in SFR and this can be seen in the distribution of grey squares that represent the full sample in all the figures. Since all figures have overlayed on these gray squares some colored contours or points, the easiest is to look at the middle figure in Fig.2 where the full sample can be best seen. By eye, I would say that the completeness limit (which is obviously above the detection limit) in SFR is above 0.01 Msun/yr and probably not very far from 0.1 Msun/yr. This naturally implies that the stellar mass range around 10^7-10^8 Msun does not provide a robust measurement of the SFR-M* main sequence and should not be used to determine the best-fitting black solid line.

    The original sample used did not have any explicit cut on mass or SFR. 
    We have assessed the completeness of the sample (including a V/V_max test)
    as well as consulting estimates from the original MPA-JHU paper (Brinchmann
    et al. 2004). They state that the sample is complete above stellar masses
    of 10^8 M_sun; we have adopted that same cutoff for all analysis in this
    paper.

    This eliminated 692 galaxies (less than 2%) star-forming galaxies from the
    sample; we have redone all analyses in the paper on this new sample. 
    The bulk of the results have not changed, and new fits and uncertainties
    are given in Table 1.

    Summary of changes:

        - arms number: 2-armed spirals no longer fall below the main sequence.
            5-armed spirals fall slightly below the main sequence, but this
            may be the result of smaller statistical samples. 
        - arm pitch angle: no change
        - bar: no change
        - mergers: slight decrease (5%) in the average SFR increase for a merging
            galaxy, from 0.307 to 0.286 dex.

    We have also put in a more explicit discussion of how the overall
    trend line is fit, including a discussion of weighting, in Section 2.

The authors do not discuss at all the completeness of the sample but this is an important information for the following reason: at low masses, i.e. in the 10^7-8 Msun range, the fraction of galaxies with an enhanced sSFR is artificially enhanced leading to an overestimate of the sSFR in this range and therefore to an underestimate of the slope of the SFR-M* relation. The fact that the slope of the SFR-M* relation is affected by this effect can be seen in Fig.1 where all colored contours in each one of the 6 panels have their barycenter and central contours systematically above the black solid line. I would expect that if the authors were to only fit the full sample in the range where the sample is complete in SFR, they would find a better agreement between their contours and the black line.

    This is confirmed; in particular, the agreement between 2-armed spirals
    and the entire SF population is now closer. The sample of 5+-armed spirals
    is the only morphology to show any offset toward lower SFR at a given mass;
    we discuss tentative reasons for this in Section 3.

    The trend for 1-armed spirals to be slightly increased in SFR per stellar
    mass is confirmed in the mass-limited sample. We interpret this as
    a signal of the presence of merging galaxies and tidal tails, as before.

As a result, we would also expect the point made in the 6th paragraph of Section 3 to be revised. In that paragraph, it is written that two-armed spiral galaxies extend slightly below the SFMS. This statement suffers in my opinion from two uncertainties. First the one that I mentioned above, i.e. the reference line is biased by a domain affected by completeness. Second, this Narms=2 subsample contains 20 times more galaxies than Narms=1 and 6 times more than Narms=3, hence before concluding that the galaxies with Narms=2 include galaxies with low sSFR and interpreting this result, the authors should first make sure that this cannot be explained by a simple statistical effect, namely that one needs a very large sample of galaxies to start sampling the low sSFR galaxies.

    The mild offset in 2-armed spirals does not appear when restricting
    the mass and SFR range of the sample. We have removed the discussion
    of this in Section 3.

In its present stage, the paper is not ready for publication since it does not provide a reliable justification of the slope of the SFR-M* relation for the whole sample of star-forming galaxies which is then used as a reference to discuss the position of various disk morphologies. We suggest that the author clearly establish a range in SFR and M* within which their galaxies fairly sample the SFR-M* relation, determine the slope of their reference trend in this range and not over the full galaxy sample including the very low SFR and M* ranges where the sample is biased due to incompleteness, and then use this new reference trend to make their comparison. If instead the authors consider their sample to fairly sample the SFR-M* relation they should explain why and explain the reason why all their subsamples are systematically biased toward higher sSFR in Fig.1. Finally, if they wish to keep their interpretation of an evidence of quenching of Narms=2 galaxies, they should discuss its statistical meaning, e.g. by randomly selecting in that sample 289 galaxies as in the Narm=1 sample and determining the % of cases where they get such low sSFR.

    As discussed above, our results are now limited to a mass range of
    10^8 < M/M_sun < 10^11.5, with star formation rates limited by the
    corresponding signal-to-noise ratios (10^-3 < SFR/(M_sun/yr) < 10^2). 
    All our analyses in this paper (plus other references in the literature
    that use this sample) indicates that this is a complete sample
    and does not suffer from the biases that would be present in a non-volume-
    limited dataset.

These changes should be very easy to make since they only concern the reference trend and they do not alter the main conclusions of the paper that rely on the comparison of the positions of each subsample, which remain unchanged e.g. in Fig.1 it will remain true that the various colored contours do not exhibit any major difference from one another. Similarly for the barred and unbarred galaxies in Fig.4 which will remain comparable.

    All figures and fits have been revised for the new mass limits.

### 2015-02-09 paper/mainsequence_morph.tex
Small changes; submitted to MNRAS round 2

\begin{abstract}
We measure the stellar mass-star formation rate relation in star-forming disk galaxies at $z<0.10$, using Galaxy~Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their average pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only boost the SFR above the main relation by $\sim0.3$~dex; this is significantly smaller than the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the various quenching mechanisms or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed, but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

The fact that star-forming galaxies live on the SFMS is one of the key observations that has been driving the development of new descriptions of how galaxies evolve \citep[eg, ][]{sch07b}. \citet{pen10a,pen12} argue that galaxies grow in stellar mass during their life as star-forming galaxies on the main sequence before having their star formation quenched either by an external mechanism (`environment quenching') or by an internal mechanism (`mass quenching').  However, \citet{del12} point out that, because of a `history bias,' galaxies of identical stellar mass may have different environmental histories that are difficult to disentangle, thus making the mass versus environment debate inherently ill-posed \citep[see also][]{van08b}. In addition, Galaxy~Zoo data has shown that the environmental dependence of galaxy star formation and that of morphology are not equivalent, mainly because of the abundance of quenched spiral galaxies, a large fraction of which are satellite galaxies \citep{ski09,bam09}. In any case, life on the main sequence appears to be governed by the action of the regulator to balance gas in- and outflows \citep{lil13}, making it a central process in galaxy evolution.

In this paper, we use data from the Sloan Digital Sky Survey \citep[SDSS;][]{yor00,str02,aba09} in combination with Galaxy~Zoo, the largest database of visual classifications of galaxy structure and morphology ever assembled \citep{lin08,lin11,wil13}, to test whether galaxy structure affects their star formation properties. We use the WMAP9 cosmology parameters of $(\Omega_m,\Omega_\Lambda,h)=(0.258,0.718,0.697)$ \citep{hin13}.

@@

Stellar masses and star formation rates are computed from optical diagnostics and taken from the MPA-JHU catalogue \citep{kau03a,bri04,sal07}. We use updated masses and activity classifications from the DR7 database.\footnote{\url{http://home.strw.leidenuniv.nl/\mytilde jarle/SDSS/}} We select only galaxies with $M_\star > 10^8 M_\odot$, for which \citet{bri04} estimate that the MPA-JHU sample is complete. Star-forming galaxies are separated from other emission-line galaxies using the standard BPT classification \citep*{bal81} below the \citet{kau03} demarcation. Galaxies classified as star-forming but with low signal-to-noise ratio $(S/N > 3)$ are also excluded. Both $M_\star$ and SFR are measured using median values extracted from the PDF. % Footnote is breaking LaTeX.

The spiral nature of the star-forming disk galaxies are identified according to the following thresholds in the GZ2 spectroscopic sample, where $p$ is the debiased vote fraction and $N$ the weighted number of total votes: $p_\textrm{features/disk} > 0.430$, $p_\textrm{not~edgeon} > 0.715$, $p_\textrm{spiral}>0.619$, and $N_\textrm{spiral}>20$. We identify sub-classes of spiral structure by weighting each galaxy according to the fraction of votes received in each morphological category.

@@

\begin{table} 
 \begin{tabular}{@{}lrcrcl}
 \hline
\multicolumn{1}{c}{Sample} &
\multicolumn{1}{c}{$N$} &
\multicolumn{1}{c}{$\alpha$} &
\multicolumn{1}{c}{$\beta$} &
\multicolumn{1}{c}{$\sigma_\alpha$} &
\multicolumn{1}{c}{$\sigma_\beta$} 
\\ 
\hline
\hline						
SF galaxies  & 48405  & $0.72$  & $-7.08$  &  $7.13\times10^{-6}$  & $6.98\times10^{-4}$  \\
\hline
1 arm        & 289    & $0.74$  & $-7.18$  &  $2.70\times10^{-5}$  & $2.69\times10^{-3}$  \\
2 arms       & 5641   & $0.78$  & $-7.69$  &  $3.70\times10^{-5}$  & $3.78\times10^{-3}$  \\
3 arms       & 997    & $0.71$  & $-6.99$  &  $4.49\times10^{-5}$  & $4.65\times10^{-3}$  \\
4 arms       & 285    & $0.71$  & $-7.05$  &  $4.80\times10^{-5}$  & $4.93\times10^{-3}$  \\
5+ arms      & 289    & $0.80$  & $-8.11$  &  $7.54\times10^{-5}$  & $7.53\times10^{-3}$  \\
can't tell   & 2017   & $0.77$  & $-7.72$  &  $4.53\times10^{-5}$  & $4.54\times10^{-3}$  \\
\hline
tight arms   & 3255   & $0.70$  & $-6.99$  &  $1.79\times10^{-3}$  & $1.86\times10^{-1}$  \\
medium arms  & 4578   & $0.77$  & $-7.62$  &  $6.81\times10^{-3}$  & $6.87\times10^{-1}$  \\
loose arms   & 1672   & $0.73$  & $-7.08$  &  $5.81\times10^{-3}$  & $5.76\times10^{-2}$  \\
\hline
barred       & 3199   & $0.75$  & $-7.38$  &  $8.53\times10^{-5}$  & $8.42\times10^{-3}$  \\
unbarred     & 11846  & $0.69$  & $-6.78$  &  $2.91\times10^{-5}$  & $2.89\times10^{-3}$  \\
\hline
merger       & 2978   & --      & $-6.73$  &  --                   & --                   \\
\hline
 \end{tabular}
\caption{Basic properties of the $M_\star-SFR$ linear fit for GZ2 star-forming galaxies. $N$ is the number of galaxies with plurality classifications for spiral arm multiplicity and pitch angles, and at a cutoff of $p_{bar}=0.4$ for barred galaxies. $\alpha$ and $\beta$ are fit according to Equation~\ref{eqn-linearfit} to data weighted by morphological vote fractions for spiral arm multiplicity and pitch angle, and to subsamples split by morphology for barred/unbarred and merging galaxies.\label{tbl-fits}}
\end{table}

To parametrize the $M_\star-SFR$ relationship for the full sample of star-forming galaxies, we apply a simple linear model for the total sample and subsamples. We apply a least-squares fit where the data are weighted by the uncertainty in SFR (computed as the mean difference in the 16th and 84th percentiles from the MPA-JHU PDFs). The data are then fit to:

@@

The sample of galaxies examined here is not explicitly constructed to be volume-limited (although \citealt{bri04} estimate that SFR and $M_\star$ are essentially complete for the mass and $S/N$ limits employed). The main reason for this is that we are comparing effects between sub-samples of galaxies using the same selection functions. One possibility is that volume-limiting would deal with galaxies in which the dust content is high enough to obscure all emission lines, even in the presence of significant star formation. Such galaxies, typically [U]LIRGs, have very low space-densities at $z<0.1$, and typically lack the regular disk structure needed to categorize it.

@@

Figure~\ref{fig-number} shows the SFR as a function of $M_\star$ for disk galaxies separated by their arm multiplicity. The GZ2 data separates disk galaxies with visible spiral arms into categories of 1, 2, 3, 4, or more than four spiral arms; there is also an additional option if the number of spiral arms cannot be accurately determined. For this analysis, galaxies in each mass/SFR bin is weighted by the vote fraction for the morphology being tested. The fit to the star-forming main sequence for all six sub-samples of spiral galaxies split by arm multiplicity tightly follows that of the total star-forming disk population. Both the slopes and offsets of each linear fit are consistent within the formal fitting errors (Table~\ref{tbl-fits}). %There are too few individual galaxies with confirmed spiral morphologies at lower masses to do a reliable fit; however, the same relation is reproduced for all arm multiplicities at $10^8<M_\star/M_\odot<10^{11}$.

@@

We have also examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in any of the results.

@@

We analyze for the first time the detailed structure of disks in large samples of star-forming galaxies as related to their position on the $M_\star$-SFR relation. This analysis is made possible by using morphological classifications from the Galaxy~Zoo~2 project. We find that neither the slope nor the dispersion of the star-forming galaxies are affected when splitting the sample into different categories of disks, including barred/unbarred galaxies, the pitch angle of spiral arms, or the number of spiral arms.

>>>>

\begin{abstract}
We measure the stellar mass-star formation rate relation in star-forming disk galaxies at $z\le0.085$, using Galaxy~Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their relative pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only boost the SFR above the main relation by $\sim0.3$~dex; this is significantly smaller than the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the various quenching mechanisms or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed, but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

The fact that star-forming galaxies live on the SFMS is one of the key observations that has been driving the development of new descriptions of how galaxies evolve \citep[eg, ][]{sch07b}. \citet{pen10a,pen12} argue that galaxies grow in stellar mass during their life as star-forming galaxies on the main sequence before having their star formation quenched either by an external mechanism (`environment quenching') or by an internal mechanism (`mass quenching').  However, \citet{del12} point out that, because of a `history bias,' galaxies of identical stellar mass may have different environmental histories that are difficult to disentangle, thus making the mass versus environment debate inherently ill-posed \citep[see also][]{van08b}. In addition, Galaxy~Zoo data has shown that the environmental dependence of galaxy star formation and that of morphology are not equivalent, mainly because of the abundance of quenched spiral galaxies, a large fraction of which are satellite galaxies \citep{ski09,bam09}. In any case, life on the main sequence appears to be governed by the action of the regulator to balance gas inflows and outflows \citep{lil13}, making the SFMS a central process in galaxy evolution.

In this paper, we use data from the Sloan Digital Sky Survey \citep[SDSS;][]{yor00,str02,aba09} in combination with Galaxy~Zoo, the largest database of visual classifications of galaxy structure and morphology ever assembled \citep{lin08,lin11,wil13}, to test whether disk structure affects a galaxy's star formation properties. We use the WMAP9 cosmology parameters of $(\Omega_m,\Omega_\Lambda,h)=(0.258,0.718,0.697)$ \citep{hin13}.

@@

Stellar masses and star formation rates are computed from optical diagnostics and taken from the MPA-JHU catalogue \citep{kau03a,bri04,sal07}. We use updated masses and activity classifications from the DR7 database.\footnote{\url{http://home.strw.leidenuniv.nl/\mytilde jarle/SDSS/}} We select only galaxies with $M_\star > 10^8 M_\odot$, for which \citet{bri04} estimate that the MPA-JHU sample is complete. Star-forming galaxies are separated from other emission-line galaxies using the standard BPT classification \citep*{bal81} below the \citet{kau03} demarcation. Galaxies classified as star-forming but with low signal-to-noise ratio $(S/N > 3)$ are also excluded. Both $M_\star$ and SFR are measured using median values extracted from the probability distribution functions.

The spiral nature of the star-forming disk galaxies is identified according to the following thresholds in the GZ2 spectroscopic sample, where $p$ is the debiased vote fraction and $N$ the weighted number of total votes: $p_\textrm{features/disk} > 0.430$, $p_\textrm{not~edgeon} > 0.715$, $p_\textrm{spiral}>0.619$, and $N_\textrm{spiral}>20$. These cuts are chosen to ensure reliable identification and with enough data points such that spiral substructure has a reasonable estimate of the associated uncertainty. Sub-classes of spiral structure are identified by weighting each galaxy according to the fraction of votes received in each morphological category.

\begin{table} 
 \begin{tabular}{@{}lrcrcl}
 \hline
\multicolumn{1}{c}{Sample} &
\multicolumn{1}{c}{$N$} &
\multicolumn{1}{c}{$\alpha$} &
\multicolumn{1}{c}{$\beta$} &
\multicolumn{1}{c}{$\sigma_\alpha$} &
\multicolumn{1}{c}{$\sigma_\beta$} 
\\ 
\hline
\hline						
SF galaxies  & 48405  & $0.72$  & $-7.08$  &  $7.13\times10^{-6}$  & $6.98\times10^{-4}$  \\
\hline
1 arm        & 289    & $0.74$  & $-7.18$  &  $2.70\times10^{-5}$  & $2.69\times10^{-3}$  \\
2 arms       & 5641   & $0.78$  & $-7.69$  &  $3.70\times10^{-5}$  & $3.78\times10^{-3}$  \\
3 arms       & 997    & $0.71$  & $-6.99$  &  $4.49\times10^{-5}$  & $4.65\times10^{-3}$  \\
4 arms       & 285    & $0.71$  & $-7.05$  &  $4.80\times10^{-5}$  & $4.93\times10^{-3}$  \\
5+ arms      & 289    & $0.80$  & $-8.11$  &  $7.54\times10^{-5}$  & $7.53\times10^{-3}$  \\
can't tell   & 2017   & $0.77$  & $-7.72$  &  $4.53\times10^{-5}$  & $4.54\times10^{-3}$  \\
\hline
tight arms   & 3255   & $0.70$  & $-6.99$  &  $1.79\times10^{-3}$  & $1.86\times10^{-1}$  \\
medium arms  & 4578   & $0.77$  & $-7.62$  &  $6.81\times10^{-3}$  & $6.87\times10^{-1}$  \\
loose arms   & 1672   & $0.73$  & $-7.08$  &  $5.81\times10^{-3}$  & $5.76\times10^{-2}$  \\
\hline
barred       & 3199   & $0.75$  & $-7.38$  &  $8.53\times10^{-5}$  & $8.42\times10^{-3}$  \\
unbarred     & 11846  & $0.69$  & $-6.78$  &  $2.91\times10^{-5}$  & $2.89\times10^{-3}$  \\
\hline
merger       & 2978   & --      & $-6.73$  &  --                   & --                   \\
\hline
 \end{tabular}
\caption{Basic properties of the $M_\star-SFR$ linear fit for GZ2 star-forming galaxies. $N$ is the number of galaxies with plurality classifications for spiral arm multiplicity and pitch angles, and at a cutoff of $p_{bar}=0.4$ for barred/unbarred galaxies. $\alpha$ and $\beta$ are fit according to Equation~\ref{eqn-linearfit} to data weighted by morphological vote fractions for spiral arm multiplicity and pitch angle, and to subsamples split by morphology for barred/unbarred and merging galaxies.\label{tbl-fits}}
\end{table}

@@

To parametrize the $M_\star-SFR$ relationship for the full sample of star-forming galaxies, we apply a simple linear model for the total sample and subsamples. We apply a least-squares fit where the data are weighted by the uncertainty in SFR (computed as the mean difference in the $16^{\rm th}$ and $84^{\rm th}$ percentiles from the MPA-JHU PDFs). The data are then fit to:

@@

The sample of galaxies examined here is not explicitly constructed to be volume-limited (although \citealt{bri04} estimate that SFR and $M_\star$ are essentially complete for the mass and $S/N$ limits employed). The main reason for this is that we are comparing effects between sub-samples of galaxies using the same selection functions. One possibility is that volume-limiting would deal with galaxies in which the dust content is high enough to obscure all emission lines, even in the presence of significant star formation. Such galaxies, typically [U]LIRGs, have very low space-densities at $z<0.1$, and typically lack the regular disk structure needed to categorize it for the morphologies considered here.

@@

Figure~\ref{fig-number} shows the SFR as a function of $M_\star$ for disk galaxies separated by their arm multiplicity. The GZ2 data separates disk galaxies with visible spiral arms into categories of 1, 2, 3, 4, or more than four spiral arms; there is also an additional option if the number of spiral arms cannot be accurately determined. For this analysis, galaxies in each mass/SFR bin is weighted by the vote fraction for the morphology being tested. The fit to the star-forming main sequence for all six sub-samples of spiral galaxies split by arm multiplicity tightly follows that of the total star-forming disk population. Both the slopes and offsets of each linear fit are consistent within the formal fitting errors in Table~\ref{tbl-fits}. %There are too few individual galaxies with confirmed spiral morphologies at lower masses to do a reliable fit; however, the same relation is reproduced for all arm multiplicities at $10^8<M_\star/M_\odot<10^{11}$.

@@

We have also examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in any of the morphologically-selected categories as compared to the general SFMS.

@@

We analyze for the first time the detailed structure of disks in large samples of star-forming galaxies in the local Universe as related to their position on the $M_\star$-SFR relation. This analysis is made possible by using morphological classifications from the Galaxy~Zoo~2 project. We find that neither the slope nor the dispersion of the star-forming galaxies are affected when splitting the sample into different categories of disks, including barred/unbarred galaxies, the pitch angle of spiral arms, or the number of spiral arms.
